---
title: "New_file"
author: "kumar"
date: "2025-03-30"
output: html_document
---
## Purpose 
This analysis project focuses on Bike uses in the New York city and client want to know the difference between the Subscribers and Non Subscribes and also wants to get strategies for converting Non subscribers into Subscribers to boost the company's revenue. He also want to know the most busiest start and end station, Which bikes are used most, peak uses hours etc.
So it is our responsibility is to present the findings to the client so that he can use then to take necessary steps to reach its goals.

## install and load the necessary packages
We will install and load the neccesary packages that are required for data cleaning, data analysis and visualization such as *tidyverse*, *dplyr* and so on. 
```{r}
# install.packages(c("readxl","dplyr","lubridate","tidyverse"))
library(readxl)
library(dplyr)
library(lubridate)
library(tidyverse)
library(ggplot2)
```

## import the .csv file
Now we will import the *citi-bike-trip-data.csv* file and check the structure of the data to know about its rows and columns
```{r}
file_path <- "C:/Users/DELL/Downloads/DA_files/Citi-Bike-Trip-Data.csv"

df <- read_csv(file_path)
head(df)
```
```{r}
# check the structure of the file
str(df)
colnames(df)
```


## Check for missing values and drop the empty column
Now we will check for any missing values in our data. Because missing values can cause the problem during the data analysis or they can lead to wrong conclusions.we can see that column **...2** is empty so we will remove it.

```{r}
# check for missing values
colSums(is.na(df))
```

## remove the *...2* column
Now we will remove the *...2* column as it is not required for our analysis or we can say that we can not clearly understand which variable do it represent.

```{r}
df <- df %>%
  select(-`...2`)

```

## Fill the missing values
We can see that in starttime column there is missing value so we will fill the missing values using median() and mode() method
```{r}
df$tripduration[is.na(df$tripduration)] <- median(df$tripduration, na.rm = TRUE)
df$starttime[is.na(df$starttime)] <- median(df$starttime, na.rm = TRUE)

get_mode <- function(v) {
  uniq_values <- unique(v)
  uniq_values[which.max(tabulate(match(v, uniq_values)))] 
}

# filling the missing values with mode
mode_user_type <- get_mode(df$usertype) 
df$usertype[is.na(df$usertype)] <- mode_user_type  

# median is used when missing values are numerical while mode is 
# used when missing values are categorical

# Verifyning if there are still any missing values.
colSums(is.na(df))

```

So clearly we can see that there are no missing values now.

## Rename the columns for better readability
We can also rename the columns for better readability such as tripduration can be renamed as Trip_duration and stoptime can be renames as Stop_Time and so on.

```{r}
colnames(df) <- c("Trip_Duration", "Start_Time", "Stop_Time","Start_Station_id","Start_station_name",
                  "Start_Station_latitude","Start_Station_longitude", "End_Station_id","End_station_name",
                  "End_Station_latitude","End_Station_longitude", "Bike_ID", "User_Type", 
                  "Birth_Year", "Gender")

colnames(df)
```

## Convert the datatypes of variables to make them of consistent format
Here we will convert the datatype of Start_Time, Stop_Time to datetime and convert User_Type column to appropriate format.

```{r}
# convert the datatype
# convert to datetime columns
df$Start_Time <- as.POSIXct(df$Start_Time, format="%Y-%m-%d %H:%M:%S")
df$Stop_Time <- as.POSIXct(df$Stop_Time, format="%Y-%m-%d %H:%M:%S")

# convert the User_Type column
df$User_Type <- as.factor(df$User_Type)
```

## Create new columns
Here we will create new column such as Start_Hour which will be made by extracting the value from the Start_Time column. Similarly Start_Day column will extracted from the Start_Time column and so on. New columns represent information such as start hour, day and month of the trip so that we can have an idea of exacttly when the trip starts.

```{r}
# create new columns
# extract hour day and month
df$Start_Hour <- hour(df$Start_Time)
df$Start_Day <- day(df$Start_Time)
df$Start_Month <- month(df$Start_Time, label = TRUE)

# trip duration in minutes
df$Trip_Duration_Minutes <- as.numeric(difftime(df$Stop_Time, df$Start_Time, units = "mins"))

```

## Remove the outliers
Now we can see that there are few columns where Start_Time is greater than the Stop_Time which is invalid because Start_Time can not be greater that the Stop_Time. So now we will remove such rows to get a cleaner data.

```{r}
df <- df %>% filter(Trip_Duration_Minutes > 0)

# verify changes
View(df)
```

## Data Analysis
Now we will analyse our data to get the insights out of it. Using these insights we can reach to a conclusion as the what the steps should be taken to reach out goals.

```{r}
# summary of data
summary(df)
```

## Most popular Start_stations
We will find the most popular stations where the trip starts.

```{r}
# most popular stations
most_papular_stations <- df %>%
  count(Start_station_name, sort = TRUE) %>%
  head(5)
print(most_papular_stations)
```

We can see that **Grove St PATH** is the most popular station where trips start.

## Find most popular end station

```{r}
# most popular end station
most_popular_end_station <- df %>%
  count(End_station_name, sort = TRUE) %>%
  head(1)
print(most_popular_end_station)
```
We can see that **Grove St PATH** is also most popular end station as well.

## Peak uses hour
Now we will find peak uses hours for cycles i.e. when cycles are in use most of the time.This will help the stakeholder to know when more cycles should be made available and we previously also calculated the most popular start and end stations so stakeeholder will know where most cycles should be made available.

```{r}
# peak uses hour
peak_uses_hour <- df %>%
  group_by(Start_Hour) %>%
  summarise(Trips = n()) %>%
  arrange(desc(Trips))
print(peak_uses_hour)
```

So we can see that 8,18 and 17 are the most busiest hours so it is recommended that more cycles should be made available during these hours.

## Bike uses
Now we will calculate that which bikes used most so that we can figure out that those bikes shold be made available in more numbers.

```{r}
# bike uses analysis
bike_uses<- df %>%
  count(Bike_ID, sort = TRUE) %>%
  head(10)
print(bike_uses)
```
We can say that bikes 26303,26229 and 26208 are most used by the riders so it is feasible that that these bikes should be made available in appropriate number in appropriate stations.

## Data visualization

Now we will visualize the insights so that audience can easily understand it.
When data is presented using visualizations such as scatter plot, pie chart, bar chart or column chart, user who do not have a data science background can also easily understand.
**Note*:* Visualization should be presented in such a way that user can understand it within 5 seconds and in next 5 seconds he should be able to extract the insights.

## Find out if user is Customer or Subscriber
Purpose of our analysis to find the ways to convert the casual riders into subscribers

```{r}
ggplot(data = df) + geom_bar(mapping = aes(x =User_Type,
  fill = User_Type)) +
  theme_minimal() +
  labs(title = "Distribution of User Types",x = 'User Type', y = 'Strength')

```
So we can see that majority of our users are Subscribers.

## Trips over time 
Now we will find out trips over a period of time and compare them

```{r}
ggplot(data = df) + geom_histogram(mapping = aes(x = Start_Time), fill = "blue",
  bins = 50, alpha = 0.6) +
  theme_minimal() +
  labs(title = "Trips Over Time", x = "Start Time", y = "Trips")

```

## Bike uses by hour and day
Now we will visualize the bike uses by hour and day to find out which days are busiest and also in that days which hours are busiest. It will give us the idea as to when the bike should be made available in large numbers.

```{r}
df %>%
  group_by(Start_Day, Start_Hour) %>%
  summarise(Trips = n()) %>%
  ggplot(aes(Start_Hour, Start_Day, fill = Trips)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Bike Usage Heatmap", x = "Hour", y = "Day")
```
Given heat map shows most busiest days and hours.

## Bike uses column chart
Given column chart shows top 10 most used bike. So these bikes should be made available at appropriate stations.So that riders can easily access these bike at those stations.
This column chart represents bikes by Bike Id metric.

```{r}
df %>%
  count(Bike_ID) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(Bike_ID, n), y = n)) +
  geom_col(fill = "purple") +
  coord_flip() +
  labs(title = "Top 10 Most Used Bikes", x = "Bike ID", y = "Number of Trips")

```

